OptiFit: a fast method for fitting amplicon sequences to existing OTUs

### 2021-02-23 paper/paper.Rmd
Rephrase _de novo_ limitations

There are three main categories of OTU clustering algorithms: closed reference,
open reference, and _de novo_ clustering. Closed reference methods assign
sequences to a set of pre-made OTUs generated from reference sequences. If a
query sequence is not within the distance threshold to any of the reference
sequences, it is discarded. While reference-based clustering is generally fast,
it is limited by the diversity represented in the reference database. Rare or
novel sequences in the sample will be lost if they are not represented in the
database. _De novo_ methods cluster sequences based on their distance to each
other, without the use of an external reference ._De novo_ clustering overcomes
the limitations of reference databases by considering only sequences in the
dataset, but is more computationally intensive and generates different OTU
assignments when new sequences are introduced.  Open reference methods take a
hybrid approach, first performing closed reference clustering, then any
sequences that cannot be assigned to reference OTUs are clustered _de novo_ to
create additional OTUs. Previous studies found that the OptiClust _de novo_
clustering algorithm created the highest quality OTU assignments based on the
Matthews correlation coefficient (MCC) [@schloss_application_2016;
@westcott_novo_2015]. However, a disadvantage of _de novo_ clustering is that
different input samples will result in different OTUs, making it difficult to
compare taxonomic composition of communities between studies or to use machine
learning models trained with _de novo_ OTUs to make predictions on new data.

To overcome the limitations of _de novo_ clustering while maintaining OTU
quality, we developed OptiFit, a reference-based clustering algorithm in the
mothur program which takes existing OTUs as the reference to fit new sequences
to. Here, we tested the OptiFit algorithm with databases or _de novo_ OTUs as
the reference and compared the performance to existing tools.

>>>>

There are three main categories of OTU clustering algorithms: closed reference,
open reference, and _de novo_ clustering. Closed reference methods assign
sequences to a set of pre-made OTUs generated from reference sequences. If a
query sequence is not within the distance threshold to any of the reference
sequences, it is discarded. While reference-based clustering is generally fast,
it is limited by the diversity represented in the reference database. Rare or
novel sequences in the sample will be lost if they are not represented in the
database. _De novo_ methods cluster sequences based on their distance to each
other, without the use of an external reference. _De novo_ clustering overcomes
the limitations of reference databases by considering only sequences in the
dataset, but is more computationally intensive and generates different OTU
assignments when new sequences are introduced. Unstable OTU assignments make it
difficult to use _de novo_ clustering to compare taxonomic composition of
communities between studies or to use machine learning models trained with _de
novo_ OTUs to make predictions on new data. Open reference methods take a hybrid
approach, first performing closed reference clustering, then any sequences that
cannot be assigned to reference OTUs are clustered _de novo_ to create
additional OTUs. Previous studies found that the OptiClust _de novo_ clustering
algorithm created the highest quality OTU assignments of all clustering methods
based on the Matthews correlation coefficient (MCC) [@schloss_application_2016;
@westcott_novo_2015].

To overcome the limitations of _de novo_ clustering while maintaining OTU
quality, we developed OptiFit, a reference-based clustering algorithm in the
mothur program which takes existing OTUs as the reference to fit new sequences
to. Here, we tested the OptiFit algorithm with the reference as databases or _de
novo_ OTUs and compared the performance to existing tools.
